Assimilating disturbance: Toward real-time monitoring and forecasting

### 2023-02-01 methods.Rmd
formatting

For the assimilation experiment, each site was assimilated separately. In the initial round of experiments the observation error variance was set to $26.5\: kg^2/m^2$ based on [@Kennedy2018] Figure 2. Post-assimilation, we then extracted a number of metrics to assess algorithm performance: the posterior probability of disturbance in 2004, the posterior biomass pool size before disturbance, after disturbance, and at the end of the time-series (2017). The difference between the predicted and observed biomass at the end of the assimilation was taken as a metric of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations. To standardize this into a unitless metric, the difference in biomass was divided by half the width of the Analysis 95% CI. Across sites the extent of filter divergence was analyzed as a function of disturbance type and post-disturbance AGB using linear models. Similarly, the ability of the algorithm to detect disturbance, represented as the mean posterior estimate of the disturbance probability, was assessed as a function of disturbance type and post-disturbance AGB using generalized linear models with a logit link and quasibinomial family.

@@

  * Score: filter divergence, disturbance detection & skill
  * Regress vs: pre-disturbance biomass, size of disturbance (abs & rel), type, etc

Divergence

As a metric to assess the prevalence of filter divergence, we calculated the difference between the last observed data point and the posterior mean of the corresponding Analysis, and then divided this by half the width of the Analysis 95% confidence interval. If the confidence interval is symmetric, then any Analysis mean that falls in the confidence interval will score between -1 and 1, while values outside this range are suggestive of filter divergence.

>>>>

For the assimilation experiment, each site was assimilated separately. In the initial round of experiments the observation error variance was set to $26.5\: kg^2/m^4$ based on [@Kennedy2018] Figure 2. Post-assimilation, we then extracted a number of metrics to assess algorithm performance: the posterior probability of disturbance in 2004, the posterior biomass pool size before disturbance, after disturbance, and at the end of the time-series (2017).
One thing we wanted to assess with the Multinomial filter is the extent of filter divergence, the situation where the model in a data assimilation algorithm becomes so confident in itself that assimilated observations have little impact on the posterior, which then diverges from observations.
As a metric to assess the prevalence of filter divergence, we calculated the difference between the last observed data point and the posterior mean of the corresponding Analysis, and then divided this by half the width of the Analysis 95% confidence interval.
If the confidence interval is symmetric, then any Analysis mean that falls in the confidence interval will score between -1 and 1, while values outside this range are suggestive of filter divergence.

Across sites the extent of filter divergence was analyzed as a function of disturbance type and post-disturbance AGB using linear models. Similarly, the ability of the algorithm to detect disturbance, represented as the mean posterior estimate of the disturbance probability, was assessed as a function of disturbance type and post-disturbance AGB using generalized linear models with a logit link and quasibinomial family.

@@

### 2023-02-01 results.Rmd
formatting

########experiment 3 - two disturb
load("exp3.Rdata") ## Generated by SingleSite_MultiDisturb_MultiPool.Rmd
pool.names=c("Cleaf","Csoil","Cwood")
i = 3
CI.Na = sapply(X = Analysis,function(y){
  x = y$MCMC
  sel.n = which(colnames(x) == paste0("X[",i,"]"))
  quantile(x[,sel.n],c(0.025,0.5,0.975))
})
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main="Multi-Disturbance")
ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
lines(CI.Na[2,],col="#b2df8a",lwd=2)
points(No[,i])
legend("bottomleft",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))

@@

##### experiment 4 - two disturb, pft switch
load("exp4.Rdata") ## SingleSite_MultiDisturb_MultiPool_MultiPFT.Rmd
pool.names=c("Cleaf","Csoil","Cwood")
i = 3
CI.Na = sapply(X = Analysis,function(y){
  x = y$MCMC
  sel.n = which(colnames(x) == paste0("X[",i,"]"))
  quantile(x[,sel.n],c(0.025,0.5,0.975))
})
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main="Multi-Disturbance, PFT switching")
ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
lines(CI.Na[2,],col="#b2df8a",lwd=2)
points(No[,i])
legend("bottomleft",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))

@@

######### experiment 5 - one disturb, 2 sites
load("exp5trim.Rdata") ##Generated by TwoSite_SingleDisturb_MultiPool.Rmd
pool.names=c("Cleaf","Csoil","Cwood")
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main="Multi-Site, Single-Disturbance")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[3]]),CI.Na[[3]][1,],CI.Na[[3]][3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[6]]),CI.Na[[6]][1,],CI.Na[[6]][3,],col=ecoforecastR::col.alpha("green",0.7))
lines(CI.Na[[3]][2,],col="#b2df8a",lwd=2)
lines(CI.Na[[6]][2,],col="green",lwd=2)
points(No[,3])
points(No[,6],pch=2)
legend("bottomleft",legend=c("Site 1","Site 2","pseudodata"),
       lwd=5,col=c("#b2df8a","green","black"))

@@

```{r divExp,echo=FALSE,fig.cap="Sensitivity of (a) filter divergence coverage, and (b) disturbance detection threshold to variance inflation of the observation error (x-axis) and process error (symbols)",fig.asp=1 , fig.width=4}
m = matrix(1:2,nrow=2,byrow = TRUE)
layout(m)
## Divergence
div.cov = sapply(stats,"[[","div.cov")
plot(Rmult[Rs],div.cov,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage",main="(a) Filter Coverage")
legend("topright",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2,title="Q")

>>>>

########experiment 3 - two disturb
load("exp3.Rdata") ## Generated by SingleSite_MultiDisturb_MultiPool.Rmd
pool.names=c("Leaf","Soil","Wood")
i = 3
CI.Na = sapply(X = Analysis,function(y){
  x = y$MCMC
  sel.n = which(colnames(x) == paste0("X[",i,"]"))
  quantile(x[,sel.n],c(0.025,0.5,0.975))
})
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main="Multi-Disturbance")
ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
lines(CI.Na[2,],col="#b2df8a",lwd=2)
points(No[,i])
legend("bottomleft",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))

@@

##### experiment 4 - two disturb, pft switch
load("exp4.Rdata") ## SingleSite_MultiDisturb_MultiPool_MultiPFT.Rmd
pool.names=c("Leaf","Soil","Wood")
i = 3
CI.Na = sapply(X = Analysis,function(y){
  x = y$MCMC
  sel.n = which(colnames(x) == paste0("X[",i,"]"))
  quantile(x[,sel.n],c(0.025,0.5,0.975))
})
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main="Multi-Disturbance, PFT switching")
ecoforecastR::ciEnvelope(1:ncol(CI.Na),CI.Na[1,],CI.Na[3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
lines(CI.Na[2,],col="#b2df8a",lwd=2)
points(No[,i])
legend("bottomleft",legend=c("Assimilation","pseudodata"),
       lwd=5,col=c("#b2df8a","black"))

@@

######### experiment 5 - one disturb, 2 sites
load("exp5trim.Rdata") ##Generated by TwoSite_SingleDisturb_MultiPool.Rmd
pool.names=c("Leaf","Soil","Wood")
rng = range(CI.Na)
if(rng[1]<1) rng[1] = 0
plot(No[,i],
     ylim = rng,type='n',
     ylab=bquote(.(pool.names[i]) ~ (kgC/m^2)),xlab="time",
     main="Multi-Site, Single-Disturbance")
  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[3]]),CI.Na[[3]][1,],CI.Na[[3]][3,],col=ecoforecastR::col.alpha('#b2df8a',0.7))
  ecoforecastR::ciEnvelope(1:ncol(CI.Na[[6]]),CI.Na[[6]][1,],CI.Na[[6]][3,],col=ecoforecastR::col.alpha("green",0.7))
lines(CI.Na[[3]][2,],col="#b2df8a",lwd=2)
lines(CI.Na[[6]][2,],col="green",lwd=2)
points(No[,3])
points(No[,6],pch=2)
legend("bottomleft",legend=c("Site 1","Site 2","pseudodata"),
       lwd=5,col=c("#b2df8a","green","black"))

@@

```{r divExp,echo=FALSE,fig.cap="Sensitivity of (a) filter divergence coverage, and (b) disturbance detection threshold to variance inflation of the observation error (x-axis) and process error (symbols)",fig.asp=1.5 , fig.width=4.5}
m = matrix(1:2,nrow=2,byrow = TRUE)
layout(m)
## Divergence
div.cov = sapply(stats,"[[","div.cov")
plot(Rmult[Rs],div.cov,col=Qs,pch=Qs,log="x",xlab="Variance Multiplier",ylab="Filter Coverage",main="(a) Filter Coverage")
legend("topright",legend=paste0(Qmult[1:2],"x"),col=1:2,pch=1:2,title="Q")
